---
title: "CLUS"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In order to work with the data I will be using the tidyverse package. It
is well designed to manipulate and work with large sets of data. The
tidy census package is designed to download census data and
automatically convert it into a tidy format for manipulation and model
building.

```{r loading packages}
pacman::p_load(tidyverse, tidycensus, writexl) #loading in and installing the packages
```

```{r Census function, message=FALSE, warning=FALSE, cache=FALSE}
# The api key is necessary in order to get the tidy census package to work.
# This url can be used to obtain a census key in about 3 min.
# http://api.census.gov/data/key_signup.html
census_api_key("4164f675c6d35b601029ed68e31ec5150dac968f")

collect <- function(year, table, geometry = FALSE){
  temp <- get_acs(geography = 'tract',
          state = "PA",
          year = year,
          table = table,
          cache_table = TRUE,
          geometry = geometry,
          output = "wide"
          ) %>% 
    rename(geoid = GEOID) %>% 
    mutate(year = year)
           
  temp %>% 
    mutate(county = str_extract(temp$NAME, "[[:alpha:]]*[[:blank:]](?=County)"),
         tract = str_remove(temp$NAME, ",.*")) %>% 
    select(geoid, year, county, tract, ends_with("E"), -NAME, -ends_with("PE")) 
}
```

First I create a function that collects the data from the American
Community Survey and changes it slightly. The default setting makes it
so every entry has 2 rows, one for the estimate of the value and its
margin of error, and one for the percentage of the estimate of the value
and its margin of error. The transformation I apply removes this second
row and just adds another 2 columns to the data.

```{r Examples, message=FALSE, cache=TRUE}
#An example of what the original function returns
test <- get_acs(geography = 'tract',
          state = "PA",
          year = 2019,
          table = "DP03",
          cache_table = TRUE,
          geometry = TRUE,
          output = "wide"
)
test %>% 
  select(GEOID, NAME, ends_with("E"), -ends_with("PE")) %>% view()
#An example of what my modified version returns
collect(2014, "DP05")
```

Here I am just running the function a bunch of times to collect all of
the versions of the data that we want from all the different years and
variables. If someone wants to optimize this code in the future it could
probably be similar with a map function or a for loop.

```{r Loading data, message=FALSE, warning=FALSE, cache=TRUE}
#Social Characteristics
SC_2019 <- collect(2019, "DP02")

#Economic characteristics
EC_2019 <- collect(2019, "DP03")

#Housing Characteristics 
HC_2019 <- collect(2019, "DP04")

#Demographic and Housing Characteristics
DC_2019 <- collect(2019, "DP05")
```

Here I am creating another function to load in some meta data about the
American Community survey that I downloaded from the census. This
enables us to get some labels about what the variables are in the tables
that we created earlier. I use some regexes to make them more readable,
however this makes some of the labels harder to understand as it removes
some clarification. If you have the census table open in a browser it
makes it much easier to understand with the combination, however more
work could be done here to make it easier to use.

```{r Creating Lables, message=FALSE, warning=FALSE}
get_labels <- function(x) {
  df <-  read_csv(x, col_names = FALSE, skip = 2) 
  
  df <- df %>% 
    slice(seq(1, as.integer(count(df)), 4))
  df <- df %>% 
    mutate(variable_c = str_remove(df$X1,"E")) %>% 
    separate(X2, into = c('f','a','b','c','d','e'), sep = "!!") %>% 
    select(-X1,-f)
}

DP02_names <- get_labels("ACSDP5Y2019.DP02_metadata_2021-03-09T174642.csv")
DP03_names <- get_labels("ACSDP5Y2019.DP03_metadata_2021-03-09T174642.csv")
DP04_names <- get_labels("ACSDP5Y2019.DP04_metadata_2021-03-09T174642.csv")
DP05_names <- get_labels("ACSDP5Y2019.DP05_metadata_2021-03-09T174642.csv")

#This line combines the labels that we created from the earlier lines of code
#into one large data set that contains all of the labels.

census_labels <- bind_rows(DP02_names, DP03_names, DP04_names, DP05_names) %>% 
  select(variable_c, a, b, c ,d, e)
```

Here we create the ACS dataset which combines all of the datasets
created earlier into one large data set, and also creates a column that
uses the labels created earlier to aid in readability of the dataset.

```{r}
municipal <- read_csv("PA_Tracts_2010_Intersect_Table.csv") 
municipal2 <- municipal%>% 
  select(GEOID10, NAMELSAD10, FID_PA_Mun) %>% 
  rename(geoid = GEOID10, tract = NAMELSAD10, municipal = FID_PA_Mun) %>% 
  mutate(geoid = as.character(geoid))

rural <- readxl::read_excel(path = "rural.xlsx")

ACS <- bind_cols(SC_2019, HC_2019, DC_2019, EC_2019) %>% 
  rename(geoid = "geoid...1",
         year = "year...2",
         county = "county...3",
         tract = "tract...4") %>% 
  select(-contains("..."), -year) %>% 
  mutate(county = str_trim(county)) %>% 
  left_join(municipal2) %>%
  left_join(rural) %>% 
  select(geoid, municipal, county, urban, tract, everything()) %>% 
  ungroup()


```

```{r}
total_population_summary <- ACS %>%
  group_by(urban) %>%
  summarise(
    total_pop = sum(DP05_0001E),
    male_prop = sum(DP05_0002E) / sum(DP05_0001E),
    female_prop = sum(DP05_0003E) / sum(DP05_0001E),
    white_prop = sum(DP05_0037E) / sum(DP05_0001E),
    black_prop = sum(DP05_0038E) / sum(DP05_0001E),
    american_indian = sum(DP05_0039E) / sum(DP05_0001E),
    asain_prop = sum(DP05_0044E) / sum(DP05_0001E),
    pacific_prop = sum(DP05_0052E) / sum(DP05_0001E)
  )

age_distribution <- ACS %>%
  group_by(urban) %>%
  summarise(
    total_pop = sum(DP05_0001E),
    'years<_5' = sum(DP05_0005E)/ sum(DP05_0001E),
    years_5_9 = sum(DP05_0006E)/ sum(DP05_0001E),
    years_10_14 = sum(DP05_0007E)/ sum(DP05_0001E),
    years_15_19 = sum(DP05_0008E)/ sum(DP05_0001E),
    years_20_24 = sum(DP05_0009E)/ sum(DP05_0001E),
    years_25_34 = sum(DP05_0010E)/ sum(DP05_0001E),
    years_35_44 = sum(DP05_0011E)/ sum(DP05_0001E),
    years_45_54 = sum(DP05_0012E)/ sum(DP05_0001E),
    years_55_59 = sum(DP05_0013E)/ sum(DP05_0001E),
    years_60_64 = sum(DP05_0014E)/ sum(DP05_0001E),
    years_65_74 = sum(DP05_0015E)/ sum(DP05_0001E),
    years_75_85 = sum(DP05_0016E)/ sum(DP05_0001E),
    'years_>85' = sum(DP05_0017E)/ sum(DP05_0001E)
  )

home_income <- ACS %>%
  group_by(urban) %>%
  summarise(
    total_households = sum(DP03_0051E),
    '<$10,000' = sum(DP03_0052E) / sum(DP03_0051E),
    '$10,000 to $14,999' = sum(DP03_0053E)/ sum(DP03_0051E),
    "$15,000 to $24,999" = sum(DP03_0054E)/ sum(DP03_0051E),
    '$25,000 to $34,999' = sum(DP03_0055E)/ sum(DP03_0051E),
    "$35,000 to $49,999" = sum(DP03_0056E)/ sum(DP03_0051E),
    "$50,000 to $74,999" = sum(DP03_0057E)/ sum(DP03_0051E),
    "$75,000 to $99,999" = sum(DP03_0058E)/ sum(DP03_0051E),
    "$100,000 to $149,999" = sum(DP03_0059E)/ sum(DP03_0051E),
    "DP03_0060  $150,000 to $199,999" = sum(DP03_0060E)/ sum(DP03_0051E),
    ">$200,000" = sum(DP03_0061E)/ sum(DP03_0051E)
  )

education_attainment <- ACS %>%
  group_by(urban) %>%
  summarise(
    population_25_and_older = sum(DP02_0059E),
    Less_than_9th_grade = sum(DP02_0060E) / sum(DP02_0059E),
    "9th_to_12th_no_diploma" = sum(DP02_0061E)/ sum(DP02_0059E),
    high_school_grad = sum(DP02_0062E)/ sum(DP02_0059E),
    some_college = sum(DP02_0063E)/ sum(DP02_0059E),
    associate = sum(DP02_0064E)/ sum(DP02_0059E),
    bachelor = sum(DP02_0065E)/ sum(DP02_0059E),
    "graduate/proffesional" = sum(DP02_0066E)/ sum(DP02_0059E),
    high_school_or_higher = sum(DP02_0067E)/ sum(DP02_0059E),
    bachelor_or_higher = sum(DP02_0068E)/ sum(DP02_0059E)
  )

ownership <- ACS %>%
  group_by(urban) %>%
  summarise(
    Occupied_housing_units = sum(DP04_0045E),
    Owner_occupied = sum(DP04_0046E) / sum(DP04_0045E),
    Renter_occupied = sum(DP04_0047E)/ sum(DP04_0045E),
    Average_household_size_of_owner_occupied_unit = median(na.omit(DP04_0048E)),
    Average_household_size_of_renter_occupied_unit = median(na.omit(DP04_0049E))
  )

housing_characteristics <- ACS %>%
  group_by(urban) %>%
  summarise(
    total_housholds = sum(DP02_0151E),
    with_computer = sum(DP02_0152E)/sum(DP02_0151E),
    with_internet = sum(DP02_0153E)/sum(DP02_0151E),
    occupied_housing_units = sum(DP04_0062E),
    utility_gas = sum(DP04_0063E)/sum(DP04_0062E),
    bottled_tanked_or_LP_gas = sum(DP04_0064E)/sum(DP04_0062E),
    electricity = sum(DP04_0065E)/sum(DP04_0062E),
    fuel_oil = sum(DP04_0066E)/sum(DP04_0062E),
    "coal/coke" = sum(DP04_0067E)/sum(DP04_0062E),
    wood = sum(DP04_0068E)/sum(DP04_0062E),
    solar = sum(DP04_0069E)/sum(DP04_0062E),
    other_fuel = sum(DP04_0070E)/sum(DP04_0062E),
    no_fuel = sum(DP04_0071E)/sum(DP04_0062E),
    lacking_plumbing = sum(DP04_0073E)/sum(DP04_0062E),
    lacking_kitchen = sum(DP04_0074E)/sum(DP04_0062E),
    lacking_telephone = sum(DP04_0075E)/sum(DP04_0062E),
    "OPR<=1" = sum(DP04_0077E)/sum(DP04_0062E),
    "OPR_1.01-1.50" = sum(DP04_0078E)/sum(DP04_0062E),
    "OPR>=1.51" = sum(DP04_0079E)/sum(DP04_0062E)
  )

housing_costs <- ACS %>%
  group_by(urban) %>%
  summarise(
    housing_units_with_mortgage = sum(DP04_0093E),
    housing_units_without_mortgage = sum(DP04_0102E),
    'housing_units_with_mortgage(excluding_SMOCAPI_NA)' = sum(DP04_0110E),
    'housing_units_without_mortgage(excluding_SMOCAPI_NA)' = sum(DP04_0117E),
    "SMOC_less_than_$500_with_mortgage" = sum(DP04_0094E)/sum(DP04_0093E),
    "SMOC_$500_to_$999_with_mortgage" = sum(DP04_0095E)/sum(DP04_0093E),
    'SMOC_$1,000_to_$1,499_with_mortgage' = sum(DP04_0096E)/sum(DP04_0093E),
    'SMOC_$1,500_to_$1,999_with_mortgage' = sum(DP04_0097E)/sum(DP04_0093E),
    'SMOC_$2,000_to_$2,499_with_mortgage '= sum(DP04_0098E)/sum(DP04_0093E),
    'SMOC_$2,500_to_$2,999_with_mortgage' = sum(DP04_0099E)/sum(DP04_0093E),
    'SMOC_$3,000_or_more_with_mortgage' = sum(DP04_0100E)/sum(DP04_0093E),
    SMOC_median_with_mortgage = median(na.omit(DP04_0101E)),
    "SMOCLess_than_$250_without_mortgage" = sum(DP04_0103E)/sum(DP04_0102E),
    "SMOC$250_to_$399_without_mortgage" = sum(DP04_0104E)/sum(DP04_0102E),
    'SMOC$400_to_$599_without_mortgage' = sum(DP04_0105E)/sum(DP04_0102E),
    'SMOC$600_to_$799_without_mortgage' = sum(DP04_0106E)/sum(DP04_0102E),
    'SMOC$800_to_$999_without_mortgage '= sum(DP04_0107E)/sum(DP04_0102E),
    'SMOC$1,000_or_more_without_mortgage' = sum(DP04_0108E)/sum(DP04_0102E),
    SMOC_median_without_mortgage = median(na.omit(DP04_0109E)),
    "SMOCAPI_with_mortgage_less_than_20.0" = sum(DP04_0111E)/sum(DP04_0110E),
    "SMOCAPI_with_mortgage_20.0_to_24.9" = sum(DP04_0112E)/sum(DP04_0110E),
    "SMOCAPI_with_mortgage_25.0_to_29.9" = sum(DP04_0113E)/sum(DP04_0110E),
    "SMOCAPI_with_mortgage_30.0_to_34.9" = sum(DP04_0114E)/sum(DP04_0110E),
    "SMOCAPI_with_mortgage_35.0_or_more" = sum(DP04_0115E)/sum(DP04_0110E),
    "SMOCAPI_with_mortgage_not_computed" = sum(DP04_0116E)/sum(DP04_0110E),
    "SMOCAPI_without_mortgage_less_than_10.0" = sum(DP04_0118E)/sum(DP04_0117E),
    "SMOCAPI_without_mortgage_10.0_to_14.9" = sum(DP04_0119E)/sum(DP04_0117E),
    "SMOCAPI_without_mortgage_15.0_to_19.9" = sum(DP04_0120E)/sum(DP04_0117E),
    "SMOCAPI_without_mortgage_20.0_to_24.9" = sum(DP04_0121E)/sum(DP04_0117E),
    "SMOCAPI_without_mortgage_25.0_to_29.9" = sum(DP04_0122E)/sum(DP04_0117E),
    "SMOCAPI_without_mortgage_30.0_to_34.9" = sum(DP04_0123E)/sum(DP04_0117E),
    "SMOCAPI_without_mortgage_35.0_or_more" = sum(DP04_0124E)/sum(DP04_0117E),
    "SMOCAPI_without_mortgage_not_computed" = sum(DP04_0125E)/sum(DP04_0117E),
  )

housing_value <- ACS %>%
  group_by(urban) %>%
  summarise(DP04_0089E)

```

```{r creating an excel worksheet, eval=FALSE, include=FALSE}


write_xlsx(
  x = list(ACS = ACS, census_labels = census_labels),
  path = "ACSv2.xlsx",
  col_names = TRUE
)

write_xlsx(
  x = list(
    total_population_summary = total_population_summary,
    age_distribution = age_distribution,
    home_income = home_income,
    education_attainment = education_attainment,
    ownership = ownership,
    housing_characteristics = housing_characteristics,
    housing_costs = housing_costs#,
    #housing_value = housing_value
  ),
  path = "ACS_summery.xlsx",
  col_names = TRUE
)
```

*Below here is just of copy past of the variable that were asked for in
the document. This is just to aid my work flow*

```{r include=FALSE, eval=FALSE}

mutate(variable_c = str_remove(df$X1,"E"),
           variable_l = str_remove(df$X2, "Estimate")) %>% 
    select(variable_c, variable_l) 
  df %>% 
    mutate(variable_l = str_replace_all(df$variable_l, "!!", "\\\n"))
  
  
ACS_long <- bind_rows(SC_2019, HC_2019, DC_2019, EC_2019) %>% 
  select(-year, -moe) %>% 
  filter(!str_detect(variable, "P$")) %>% 
  rename(variable_c = variable)
  

test <- left_join(ACS_long, census_labels, by = "variable_c")
  
test1 <- test %>% pivot_wider(id_cols = c(geoid, county, tract), names_from = c(variable_c, variable_l), values_from = estimate) 



  
```

summery stats a. total population estimates for each municipality and
sex and race ratio b. age distribution c. family income distribution d.
educational attainment distribution e. homeownership rate vs renter rate
f. distribution of housing characteristics (percentages of lacking
complete plumbing, lacking complete kitchen facilities, no telephone
service, without a computer, without internet subscription, types of
heating fuels used, and average number of occupants per room) g. medians
of monthly housing cost and monthly housing costs as a percentage of
household income (run this for homeowners only) h. median property
value, ratio of value to household income and real estate taxes (for
homeowners only)

Selected housing characteristics: housing occupancy (occupied housing
units, vacant housing units, homeowner vacancy rate, rental vacancy
rate); Units in structure; year structure built; \# of rooms; \#of
bedrooms; housing tenure (owner vs. renter); year householder moved in;
house heating fuel; plumbing, kitchen, and telephone service; occupants
per room; mortgage status;

Race and ethnicity (when downloading the table, keep estimate and
percent) Population and people (age and sex) Selected social
characteristics (household by type, marital status, educational
attainment, disability status, U.S. citizenship status and year of
entry, computers and internet use) Financial characteristics (household
income, monthly housing costs, monthly housing costs as a percentage of
household income, mortgage status, ratio of value to household income,
real estate taxes) idea storage
